Harmony: A Verifiable Systems Programming Language

### 2022-01-14 paper/paper.tex
Changing the spin

\title{Concurrent and Distributed Programming in Harmony}
\author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
\date{Department of Computer Science, Cornell University}

@@

\emph{Harmony} is a Python-like programming language designed to teach
concurrent and distributed programming to students who do not have a
background in formal methods.  Harmony programs are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\end{compactitem}

@@

Harmony programs have three sources of non-determinism:
<{choose}> operations that select a value from a set,
\emph{interleaving} due to multi-threading, and \emph{interrupts}.
\autoref{fig:helloworld} demonstrates the <{choose}> and the <{spawn}>
statements.  Since Harmony programs are model-checked and multiple
executions are explored across a finite set of states, the output of a
Harmony program is a finite automaton that describes all possible
outputs of the program.  Groups of statements can be made atomic
using the <{atomically}> keyword.  Finally, Harmony supports
waiting until some condition is reached with the <{when}> statement.

@@

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level programming language, the ability to concisely
specify an interface, the ability to compare behaviors of specification
and implementation, the ability to modularly model check programs,
and support for continuations and interrupts.  Harmony finds various common
problems automatically without a complicated property specification
language, such as Harmony's ability to find instances of undesirable
busy waiting.

This paper provides an overview of the Harmony language and Charm,
its model checker, and how they may be used by programmers to
develop code for reliable concurrent or distributed applications.

@@

<{:end:}>
\end{tcolorbox}
}
\\
(a) & (b)
\end{tabular}
\end{center}
\caption{Lock specification and a spinlock implementation based on
test-and-set}
\label{fig:synch}
\end{figure*}

@@

>>>>

\title{Harmony: A Verifiable Systems Programming Language}
\author{Robbert van Renesse \and William Ma \and Kevin Sun \and Anthony Yang}
\date{Department of Computer Science, Cornell University}

@@

\emph{Harmony} is a Python-like programming language designed to teach
systems programming and verification techniques to students who do not have a
background in formal methods.  Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\end{compactitem}

@@

Harmony programs have three sources of non-determinism:
<{choose}> operations that select a value from a set,
\emph{interleaving} due to multi-threading, and \emph{interrupts}.
\autoref{fig:helloworld} demonstrates the <{choose}> and the <{spawn}>
statements.  Since Harmony programs are model-checked and multiple
executions are explored across a finite set of states, the output of a
Harmony program is a finite automaton that describes all possible
outputs of the program.  Groups of statements can be made atomic
using the <{atomically}> keyword.  Harmony supports
waiting until some condition is reached with its <{when}> statement.

@@

Harmony supports continuations as first-class objects, allowing the
state of a thread can be saved in a variable and later restored.
Continuations enable the implementation of specialized thread
schedulers as well as the implementation of thread forking by
reinstantiating a saved thread state multiple times.

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level programming language, the ability to concisely
specify an interface, the ability to compare behaviors of specification
and implementation, the ability to modularly model check programs,
and support for continuations and interrupts.  In addition, Harmony
finds various common problems automatically without a complicated
property specification language, such as Harmony's ability to find
instances of undesirable busy waiting.

This paper provides an overview of the Harmony language and Charm,
its model checker, and how they may be used by programmers to
develop specify and implement reliable concurrent and/or
distributed systems.

@@

<{:end:}>
\end{tcolorbox}
}
\\
(a) & (b)
\end{tabular}
\end{center}
\caption{(a) Lock specification and (b) a spinlock implementation based on
test-and-set}
\label{fig:synch}
\end{figure*}

@@

Dist\-Algo~\cite{DistAlgo} is a programming language based on Python to
simplify the specification and verification of distributed algorithms.
Dist\-Algo includes a framework for runtime checking of safety and liveness
properties~\cite{LS20}.
